---
title: "lasso_model"
author: "Kate Weaver"
date: "1/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, load_packages, warning=FALSE, message=FALSE}
library(SummarizedExperiment)
library(rtracklayer)
library(tidyverse)
library(data.table)
library(readxl)
library(janitor)
library(ggplot2)
library(AnnotationDbi)
library(EnsDb.Hsapiens.v86)
library(glmmLasso)
library(edgeR)
library(caret)
library(recipes)
library(ROCR)
```

importing SummarizedExperiment Data and transforming to raw counts data
```{r load_data_se}
#load("~/Dropbox/projects/2022_euploid_lasso/create_summarized_experiment_allNov2021.Rdata")
load("~/mccoyLab/collabs/doubleseq_2021/summarized_experiment/create_summarized_experiment_allNov2021.Rdata")
counts <- as.matrix(assays(seAll)$counts)
for (col in 1:ncol(counts)){
  colnames(counts)[col]<-strsplit(sub("Aligned.sortedByCoord.out.bam", "", colnames(counts)[col]), "_")[[1]][1]
}
```

```{r, load_andsubsetby_sets}
#subset to just training data/samples
#embryoID_by_set <- read.csv("~/Dropbox/projects/2022_euploid_lasso/embryo_bySet_full_kw_20211217.csv", col.names=c("embryoID", "set"))
embryoID_by_set <- read.csv("~/mccoyLab/collabs/doubleseq_2021/data_split/embryo_bySet_full_kw_20211217.csv", col.names=c("embryoID", "set"))
train_embryoIDs = embryoID_by_set$embryoID[embryoID_by_set$set == "train"]
```

```{r load_andsubsetby_gencode}
#subset to genes with matchind gencode ensembl gene IDs on chr1-22
#file_gencode <- "~/Dropbox/projects/2022_euploid_lasso/gencode.v34.annotation.gtf"
file_gencode <- "~/genomes/hg38_genome/gencode.v34.annotation.gtf"

gtf <- rtracklayer::import(file_gencode) %>% 
  as.data.frame() %>% 
  dplyr::filter(type == "gene") %>% 
  dplyr::select(gene_id, seqnames, width, gene_name) %>% 
  dplyr::rename(ensembl_gene_id = gene_id) %>% 
  dplyr::rename(chromosome_name = seqnames) %>% 
  dplyr::rename(gene_name = gene_name) %>% 
  dplyr::rename(length = width)

gene_table <- gtf[match(rownames(counts), gtf$ensembl_gene_id),]
gene_table <- gene_table[gene_table$chromosome_name %in% paste0("chr", 1:22),]
counts <- counts[gene_table$ensembl_gene_id,]
```

```{r filter_expression}
#Use more stringent filtering
dgeFullData <- DGEList(counts)
#normalize counts by TMM
TMMFullData <- calcNormFactors(dgeFullData, method="TMM")

#TPM Calculation
calc_tpm <- function(x, gene.length) {
  x <- as.matrix(x)
  len.norm.lib.size <- colSums(x / gene.length)
  return((t(t(x) / len.norm.lib.size) * 1e06)/ gene.length)
}

#creates a matrix with calculated TPM values for each sample from TMM normalized counts and the gene lengths
rawTPMvals <- calc_tpm(TMMFullData, gene.length = gene_table$length)
```

select genes from any of difexpression analyses
```{r downselect_genes}
#design0_res <- read.csv("~/Dropbox/projects/2022_euploid_lasso/results_trainSPA_onlyBatch.qval.csv")
design0_res <- read.csv("~/mccoyLab/collabs/doubleseq_2021/dif_expression_results/results_trainSPA_onlyBatch.qval.csv")
#design3_res <- read.csv("~/Dropbox/projects/2022_euploid_lasso/results_trainSPA_dupCor_withBatch.qval.csv")
design3_res <- read.csv("~/mccoyLab/collabs/doubleseq_2021/dif_expression_results/results_trainSPA_dupCor_withBatch.qval.csv")
#featselect_res <- read.csv("~/Dropbox/projects/2022_euploid_lasso/results_trainSPA_featselect_withBatch.qval.csv")
featselect_res <- read.csv("~/mccoyLab/collabs/doubleseq_2021/dif_expression_results/results_trainSPA_featselect_withBatch.qval.csv")
genes_oi <- full_join(full_join(featselect_res, design3_res), design0_res)

counts_train <- as.matrix(t(as.data.frame(t(cbind(rawTPMvals[genes_oi$X,])))[train_embryoIDs,])) #64 genes, 107 samples
```

```{r load_andmatchby_meta}
#Importing metadata to differentiate between pregnant and not pregnant and other main covariates
#meta <- read.csv("~/Dropbox/projects/2022_euploid_lasso/tidied_meta_CREATE_kw_20211217.csv", row.names = 1) %>%
meta <- read.csv("~/mccoyLab/collabs/doubleseq_2021/tidied_meta/tidied_meta_CREATE_kw_20211217.csv", row.names = 1) %>% 
  as.data.frame() %>% 
  mutate(across(c("AOD", "GC", "Infertility_type", "Previous_pregnancy", "Past_surgical_hist", "Pregnant", "Ongoing_pregnancy", "Final_outcome", "Embryo_grade_at_freezing", "Interpretation", "cDNA_RT_Date", "Library_Prep_Date", "Sequencing_Date", "Study_Participant_ID"), as.factor)) %>%
  mutate(across(c("InfD_SSM_GC", "InfD_Egg_factor", "InfD_MF", "InfD_Uterine_factor", "InfD_TF", "InfD_RPL", "InfD_RIF", "InfD_Unexplained", "PMdH_none", "PMdH_vasculitis", "PMdH_immune", "PMdH_stress_hormones"), as.factor))

#subsetting embryos to include only ones with sequencing data 
remove <- setdiff(train_embryoIDs, colnames(counts))
if (length(which(rownames(meta) %in%  remove)) > 0){
  train_embryoIDs <- train_embryoIDs[-which(train_embryoIDs %in% remove)]
}

#subsetting meta to just the training rows            
meta_train <- meta[train_embryoIDs,]
```

select only relevant metadata columns
```{r subsetmeta}
columns_to_remove <- c("Ongoing_pregnancy", "Final_outcome", #irrelevant outcomes
                       "Time_to_OPU_hr", "Time_to_stripping_hr", "Time_to_ICSI_hr", "Time_to_Biopsy_hr", "AMH", #NA values
                       "Sildenafil") #only one factor level
to_remove <- which(colnames(meta_train) %in% columns_to_remove)


meta_covs <- meta_train[,-to_remove]
meta_covs$Pregnant=meta_covs$Pregnant==1
```


make a full dataframe with sample as row and covariate variable or gene as column
```{r onedf}
df <- merge(meta_covs, as.data.frame(t(counts_train)), by=0) %>% `rownames<-`(.[,"Row.names"]) %>% dplyr::select(-Row.names) %>% 
  as.data.frame() %>%
  dplyr::mutate(across(c(Study_Participant_ID, cDNA_RT_Date, Library_Prep_Date, Sequencing_Date, Transferring_physician, Transfer_catheter_used), as.character)) %>%
  dplyr::mutate(across(c(Study_Participant_ID, cDNA_RT_Date, Library_Prep_Date, Sequencing_Date, Transferring_physician, Transfer_catheter_used), as.factor)) %>%
  dplyr::mutate_at(vars(starts_with("ENSG")), as.numeric)

dim(df)  #107 samples; 117 variables
```

load test dataframe
```{r load_test_data}

test_embryoIDs = embryoID_by_set$embryoID[embryoID_by_set$set == "test"]

#subsetting embryos to include only ones with sequencing data 
remove <- setdiff(test_embryoIDs, colnames(counts))
if (length(which(rownames(meta) %in%  remove)) > 0){
  test_embryoIDs <- test_embryoIDs[-which(test_embryoIDs %in% remove)]
}

#subsetting meta and counts data to just the training rows            
meta_test <- meta[test_embryoIDs, -to_remove]
counts_test <- as.matrix(t(as.data.frame(t(cbind(rawTPMvals[genes_oi$X,])))[test_embryoIDs,])) #64 genes, 36 samples

meta_test$Pregnant=meta_test$Pregnant==1

df_test <- merge(meta_test, as.data.frame(t(counts_test)), by=0) %>% 
  `rownames<-`(.[,"Row.names"]) %>% dplyr::select(-Row.names) %>% 
  as.data.frame() %>%
  dplyr::mutate(across(c(Study_Participant_ID, cDNA_RT_Date, Library_Prep_Date, Sequencing_Date, Transferring_physician, Transfer_catheter_used), as.character)) %>%
  dplyr::mutate(across(c(Study_Participant_ID, cDNA_RT_Date, Library_Prep_Date, Sequencing_Date, Transferring_physician, Transfer_catheter_used), as.factor)) %>%
  dplyr::mutate_at(vars(starts_with("ENSG")), as.numeric) 

dim(df_test) #36 samples; 117 variables
```

```{r, lassomodel}
rnd_form <- list("Study_Participant_ID"=~1, "cDNA_RT_Date"=~1, "Library_Prep_Date"=~1,"Sequencing_Date"=~1, "Transferring_physician"=~1, "Transfer_catheter_used"=~1)
fix_form <- reformulate(setdiff(colnames(df), c("Pregnant", "Study_Participant_ID", "cDNA_RT_Date", "Library_Prep_Date", "Sequencing_Date","Transferring_physician", "Transfer_catheter_used")), response="Pregnant")
fix_form
```

scale data 
```{r, scale}
scale_function <- recipe(~ ., df) %>%
  step_normalize(all_numeric()) %>%
  prep()

df_train_scaled <- scale_function %>% 
  bake(df, all_numeric())
df_train_nn <- select_if(df, negate(is.numeric))
df_train_scaled <- cbind(df_train_nn, df_train_scaled)

df_test_scaled <- scale_function %>% 
  bake(df_test, all_numeric())
df_test_nn <- select_if(df_test, negate(is.numeric))
df_test_scaled <- cbind(df_test_nn, df_test_scaled)



```


had to copy formula from above and add the as.factor's to each of the categorical predictor variables :(
```{r, lassomodel_pt2, warning=FALSE}
lambda <- seq(500, 0, by = -5)
bic_vec <- rep(Inf, length(lambda))
m_list <- list()

for(j in 1:length(lambda)){
  lassomod <- try(glmmLasso(fix = Pregnant ~ 1 + as.factor(AOD) + 
                                  as.factor(GC) +
                                  Oocyte_Age + Uterus_Age + Sperm_Age + BMI + 
                                  as.factor(Infertility_type) +
                                  as.factor(Previous_pregnancy) +
                                  as.factor(Past_surgical_hist) + 
                                  FSH_D3 + AFC + LH + E2_Day_2_3 +
                                  as.factor(Type_of_trigger) + 
                                  Total_Gn + 
                                  E2_on_trigger + Num_eggs + Num_MII + Num_2PN + Num_Blasts + TMC + 
                                  as.factor(Embryo_grade_at_freezing) +
                                  as.factor(Interpretation) + 
                                  as.factor(Natural_cycle) + 
                                  as.factor(Medication) + 
                                  lining_thickness_mm + 
                                  as.factor(Prednisone) + 
                                  as.factor(Fragmin) + 
                                  as.factor(EG) + 
                                  as.factor(HCG) + 
                                  as.factor(IL) + 
                                  as.factor(GCSF) + 
                                  as.factor(Metformin) + 
                                  as.factor(PRP) + 
                                  as.factor(InfD_SSM_GC) + 
                                  as.factor(InfD_Egg_factor) + 
                                  as.factor(InfD_MF) + 
                                  as.factor(InfD_Uterine_factor) + 
                                  as.factor(InfD_TF) + 
                                  as.factor(InfD_RPL) + 
                                  as.factor(InfD_RIF) + 
                                  as.factor(InfD_Unexplained) + 
                                  as.factor(PMdH_none) + 
                                  as.factor(PMdH_vasculitis) + 
                                  as.factor(PMdH_immune) + 
                                  as.factor(PMdH_stress_hormones) + 
                                  ENSG00000139908.15 + ENSG00000198526.7 + 
                                  ENSG00000260876.6 + ENSG00000274979.1 + ENSG00000235523.1 + 
                                  ENSG00000158220.14 + ENSG00000197496.6 + ENSG00000143297.19 + 
                                  ENSG00000135355.4 + ENSG00000276754.1 + ENSG00000139531.13 + 
                                  ENSG00000197651.4 + ENSG00000159882.13 + ENSG00000287462.1 + 
                                  ENSG00000285696.1 + ENSG00000243742.5 + ENSG00000163689.20 + 
                                  ENSG00000184992.13 + ENSG00000228889.7 + ENSG00000167419.11 + 
                                  ENSG00000234277.3 + ENSG00000286963.1 + ENSG00000139438.6 + 
                                  ENSG00000251244.2 + ENSG00000170236.16 + ENSG00000164167.12 + 
                                  ENSG00000186723.4 + ENSG00000170949.17 + ENSG00000235713.1 + 
                                  ENSG00000140265.12 + ENSG00000151445.16 + ENSG00000171243.8 + 
                                  ENSG00000214894.6 + ENSG00000164338.10 + ENSG00000186417.14 + 
                                  ENSG00000233208.6 + ENSG00000260698.1 + ENSG00000253550.1 + 
                                  ENSG00000187720.14 + ENSG00000163746.11 + ENSG00000254180.2 + 
                                  ENSG00000198440.9 + ENSG00000279315.1 + ENSG00000118292.9 + 
                                  ENSG00000117281.16 + ENSG00000138496.16 + ENSG00000283069.1 + 
                                  ENSG00000131059.12 + ENSG00000230299.1 + ENSG00000223522.2 + 
                                  ENSG00000206043.7 + ENSG00000280739.3 + ENSG00000164707.15 + 
                                  ENSG00000259683.1 + ENSG00000118271.11 + ENSG00000090612.22 + 
                                  ENSG00000075239.14 + ENSG00000112984.12 + ENSG00000065308.5 + 
                                  ENSG00000235275.3 + ENSG00000120875.9 + ENSG00000160570.14 + 
                                  ENSG00000160570.14.1 + ENSG00000054965.10, 
                            rnd = rnd_form,
                            data = df_train_scaled,
                            lambda = lambda[j],
                            family = binomial(),
                            switch.NR = FALSE,
                            final.re = TRUE))

  if(class(lassomod)!="try-error"){ 
    bic_vec[j] <- lassomod$bic
    m_list[[j]] <- lassomod
  }
}
min_ind <- which.min(bic_vec)
opt_m <- m_list[[min_ind]]
summary(opt_m)
lambda[min_ind]
bic_vec[min_ind]
```

```{r, predict_test}

df_test_scaled$pred_implant_prob <- predict(opt_m, newdata=df_test_scaled)
```

```{r visualizations}
conf_mat <- caret::confusionMatrix(data = as.factor(df_test_scaled$pred_implant_prob > 0.5), 
                reference = as.factor(df_test_scaled$Pregnant == 1))
conf_mat
conf_mat_df <- data.frame(conf_mat$table)
conf_mat_df$relfreq <- conf_mat_df$Freq/sum(conf_mat_df$Freq)
cm_plt <- ggplot(conf_mat_df, aes(x = Reference, y = Prediction, fill = relfreq)) + 
          geom_tile() +
          geom_text(aes(label = Freq), vjust = .5, fontface  = "bold", alpha = 1) +
          theme_bw() +
          xlim(rev(levels(conf_mat_df$Reference))) + ggtitle("Confusion Matrix: Implantation Success")
cm_plt

pred <- prediction(df_test_scaled$pred_implant_prob, as.factor(df_test_scaled$Pregnant == 1))    
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
df_plot <- data.frame(Curve=1, 
                 FalsePositive=c(perf@x.values[[1]]),
                 TruePositive=c(perf@y.values[[1]]))
plt <- ggplot(df_plot, aes(x=FalsePositive, y=TruePositive, color=as.factor(Curve))) + geom_line() + geom_abline(mapping = NULL, 0,1, linetype=3) + xlab("False Positive Rate") + ylab("True Positive Rate") +ggtitle("ROC Curve: Implantation Success") + theme_bw() + theme(legend.position = "none") + theme(panel.background = element_blank(), panel.grid = element_blank())
plt
auc <- performance(pred, "auc")
auc@y.values[[1]]
```

### plot for training data too

```{r predict_train}
df_train_scaled$pred_implant_prob <- predict(opt_m, newdata=df_train_scaled)
```

```{r visualizations_train}
conf_mat_train <- caret::confusionMatrix(data = as.factor(df_train_scaled$pred_implant_prob > 0.5), 
                reference = as.factor(df_train_scaled$Pregnant == 1))
conf_mat_train
conf_mat_train_df <- data.frame(conf_mat_train$table)
conf_mat_train_df$relfreq <- conf_mat_train_df$Freq/sum(conf_mat_train_df$Freq)
cm_train_plt <- ggplot(conf_mat_train_df, aes(x = Reference, y = Prediction, fill = relfreq)) + 
          geom_tile() +
          geom_text(aes(label = Freq), vjust = .5, fontface  = "bold", alpha = 1) +
          theme_bw() +
          xlim(rev(levels(conf_mat_train_df$Reference))) + ggtitle("Confusion Matrix: Implantation Success -- Training Data")
cm_train_plt

pred_train <- prediction(df_train_scaled$pred_implant_prob, as.factor(df_train_scaled$Pregnant == 1))    
perf_train <- performance(pred_train, measure = "tpr", x.measure = "fpr")
df_plot_train <- data.frame(Curve=1, 
                 FalsePositive=c(perf_train@x.values[[1]]),
                 TruePositive=c(perf_train@y.values[[1]]))
plt_train <- ggplot(df_plot_train, aes(x=FalsePositive, y=TruePositive, color=as.factor(Curve))) + geom_line() + geom_abline(mapping = NULL, 0,1, linetype=3) + xlab("False Positive Rate") + ylab("True Positive Rate") +ggtitle("ROC Curve: Implantation Success -- Training Data") + theme_bw() + theme(legend.position = "none") + theme(panel.background = element_blank(), panel.grid = element_blank())
plt_train
auc_train <- performance(pred_train, "auc")
auc_train@y.values[[1]]
```
We're likely under regularizing then